\chapter*{Acknowledgements}

\vspace{-0.3cm}

{
\setstretch{1.0}
I could not have completed this book without the support of my wife, Marie. She proofread every word in this book and gave valuable input on its content. I want to thank my parents for their encouragement of my curiosity and scientific endeavors since childhood. My sons, William and Vincent, make my life and research more joyful, especially when I see they are curious and enthusiastic to see me testing antennas, receivers, and hardware.

Navigating through the mountain of information in different ICAO documents is complicated. I am grateful to have met Huy V\^u and supervised his master thesis project. He was able to find almost any needed information regarding Mode~S, which was a great help for this book and the pyModeS library. He is now a data analyst at \emph{LVNL}, the Dutch air traffic control, and we still work together on interesting research topics. 

In the summer of 2017, I received an email from a researcher from \emph{ONERA}, the French aerospace research institute, who wanted to meet me for a coffee to discuss Mode S. That coffee discussion with Xavier Olive has turned into a fruitful collaboration. I want to say thanks to Xavier for his input on Mode S, ADS-B, and all other related and unrelated discussions.

Before the notion of this book ever existed, I was a new PhD student at TU Delft's CNS/ATM research group. I would like to thank Jacco Hoekstra and Joost Ellerbroek, who were my promotors and provided me with great support for all my PhD research topics. I am also extremely happy to have been able to continue working with them as a colleague since 2019. I would also like to extend my thanks to the Dean of our faculty, Henri Werij, for his consistent support and for offering me the opportunity to continue my research at the faculty. 

My research often relies on Mode S and ADS-B data from regions beyond the coverage of our antenna in Delft. This is only possible with the help of crowd-sourced networks. I want to thank Martin Strohmeier from \emph{OpenSky network}, Sean Atkinson from \emph{FlightRadar24}, and James Stanford from \emph{ADS-B Exchange} for sharing their data with me for different research projects over the past years.

The manual was an open-access book from the beginning. I heartily appreciate all the GitHub contributors who made suggestions and edits for this book and pyModeS library.\footnote{The most up-to-date list of contributors can be found at: \\https://github.com/junzis/the-1090mhz-riddle/graphs/contributors \\https://github.com/junzis/pyModeS/graphs/contributors} Since I decided to publish this book with TU Delft OPEN Publishing, our publishing officer, Frederique Belliard, has been extremely helpful in making this book a reality. Finally, I would also like to thank Petr Jonas, Xavier Olive, and Enrico Spinielli for their comprehensive peer reviews that led to the successful completion of this book.
}